Reagents: expressing and composing fine-grained concurrency

### 2011-11-06 paper/paper.tex
work on implementation writing

We test this principle empirically in~\secref{performance} by comparing multiple
reagent-based collections to their hand-written counterparts, as well as to
lock-based and STM-based implementations.  The benchmarks include both single
operations used in isolation (where little overhead for reagentes is expected)
and combined atomic transfers.  We compare at both high and low levels of
contention.  Reagents perform universally better than the lock- and STM-based
implementations, and are indeed competetive with hand-written lock-free
implementations.

@@

We noted that \lstinline{s.push} can only fail due to a failed CAS.
Such a failure is \emph{transient}: it can only be caused by active
interference from another thread.  A reagent that has failed
transiently should retry, rather than block, following the concurrency
patterns laid out in \secref{background}.  On the other hand,
\lstinline{s.pop} can fail in a \emph{permanent} way: the underlying
stack \lstinline{s} may be empty.  This failure is permanent in the
sense that only activity by another thread can enable the reagent to
proceed.  When faced with a permanent failure, a reagent should block
until signalled that the underlying state has changed.

@@

Here we begin to see the benefits of the reagent abstraction.  Both of
the example combinations work regardless of how the underlying stacks
are implemented.  If both stacks use elimination backoff, the
conjoined operations will potentially use elimination on both
simultaneously.  This behavior is entirely emergent; it does not
require any code on the part of the stack author, and it does not
require the stack user to know anything about the implementation.
Reagents can be composed in unanticpated ways.

@@

In other words, \lstinline{tryReact} takes the input (type \lstinline{A}) to the
reagent and the reaction built up so far, and either completes the reaction,
returning a result (type \listinline{B}), or fails, returning one of the
failure singletons (\lstinline{Block} or \lstinline{Retry}).

@@

shown in \figref{reaction}

@@

\begin{figure}
\begin{lstlisting}
def !(a: A): B = {
  val backoff = new Backoff
  def retryLoop(shouldBlock: Boolean): B = {
    if (maySync || shouldBlock) {
      val offer = new Offer[B]
      tryReact(a, empty, offer) match {
        case Block => 
          LockSupport.park
          if (offer.rescind) retryLoop(true)
          else offer.answer
        case Retry =>
          backoff.once
          if (offer.rescind) retryLoop(maySync)
          else offer.answer
      }
    } else {
      tryReact(a, empty, null) match {
        case Block => retryLoop(true)
        case Retry => backoff.once; retryLoop(maySync)
        case ans   => ans.asInstanceOf[B]
      }
    }
  retryLoop(false)
}
\end{lstlisting}
\nocaptionrule
\caption{The \lstinline{!} method, defined in \lstinline{Reagent[A,B]}}
\label{fig:reaction}
\end{figure}
%TODO: note tail recursion

@@

@@

kcas implementations: \cite{Fraser2007,Luchangco2003,Attiya2008}

\begin{lstlisting}
def tryReact(u: Unit, rx: Reaction, offer: Offer[B]) = 
  if (!rx.hasCAS && !k.hasCAS) {
    if (ref.cas(oldV, newV))
      k.tryReact((), rx, offer)
    else Retry
  } else {
    k.tryReact((), rx.withCAS(ref, oldV, newV), offer)
  }
\end{lstlisting}

@@

\begin{lstlisting}
def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
  r1.tryReact(a, rx, offer) match {
    case Retry => 
      r2.tryReact(a, rx, offer) match {
        case Retry => Retry
        case Block => Retry // must still retry r1
        case ans   => ans
      }
    case Block => r2.tryReact(a, rx, offer)
    case ans => ans
  }
\end{lstlisting}

@@

\begin{lstlisting}
  def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any = 
    c(a).tryReact((), rx, offer)
\end{lstlisting}

>>>>

We test this principle empirically in~\secref{performance} by comparing multiple
reagent-based collections to their hand-written counterparts, as well as to
lock-based and STM-based implementations.  The benchmarks include both single
operations used in isolation (where little overhead for reagents is expected)
and combined atomic transfers.  We compare at both high and low levels of
contention.  Reagents perform universally better than the lock- and STM-based
implementations, and are indeed competetive with hand-written lock-free
implementations.

@@

We noted that \lstinline{s.push} can only fail due to a failed CAS.
Such a failure is \emph{transient}: it can only be caused by active
interference from another thread.  A reagent that has failed
transiently should retry, rather than block, following the concurrency
patterns laid out in \secref{background}.  On the other hand,
\lstinline{s.pop} can fail in a \emph{permanent} way: the underlying
stack \lstinline{s} may be empty.  This failure is permanent in the
sense that only activity by another thread can enable the reagent to
proceed.  When faced with a permanent failure, a reagent should block
until signaled that the underlying state has changed.

@@

Here we begin to see the benefits of the reagent abstraction.  Both of
the example combinations work regardless of how the underlying stacks
are implemented.  If both stacks use elimination backoff, the
conjoined operations will potentially use elimination on both
simultaneously.  This behavior is entirely emergent; it does not
require any code on the part of the stack author, and it does not
require the stack user to know anything about the implementation.
Reagents can be composed in unanticipated ways.

@@

In other words, \lstinline{tryReact} takes the input (type \lstinline{A}) to
the reagent and the reaction built up so far, and either completes the
reaction, returning a result (type \listinline{B}), or fails, returning one of
the failure singletons (\lstinline{Block} or \lstinline{Retry}).  The
remaining argument, \lstinline{offer}, is used for synchronization and
communication between reagents, which we explain next.

\subsection{Offers, communication and continuations}
\label{sec:offers}

@@

@@

\begin{figure}
\begin{lstlisting}
def !(a: A): B = {
  val backoff = new Backoff
  def withoutOffer(): B =
    tryReact(a, empty, null) match {
      case Block => withOffer()
      case Retry => 
        backoff.once()
        if (maySync) withOffer() else withoutOffer()
      case ans   => ans.asInstanceOf[B]
    }
  def withOffer(): B = {
    val offer = new Offer[B]
    tryReact(a, empty, offer) match {
      case (f: Failure) => 
        if (f == Block) park() else backoff.once()
        if (offer.rescind) withOffer() else offer.answer
      case ans => ans.asInstanceOf[B]
    }
  }
  withoutOffer()
}
\end{lstlisting}
\nocaptionrule
\caption{The \lstinline{!} method, defined in \lstinline{Reagent[A,B]}}
\label{fig:reaction}
\end{figure}
%TODO: note tail recursion

The code for performing a reaction is given in the \lstinline{!} method
definition for \lstinline{Reagent[A,B]}, shown in~\figref{reaction}.  This
method provides two generalized versions of the optimistic retry loops we
described in~\secref{background}.  The retry loops are written as a local,
tail-recursive functions, which Scala compiles down to loops.

The first retry loop, \lstinline{withoutOffer}, attempts to perform the
reaction without making visible offers to other reagents.  It may, however,
find and consume offers from other reagents as necessary for message passing.
To initiate the reaction, \lstinline{withoutOffer} calls the abstract
\lstinline{tryReact} method with the input \lstinline{a}, an empty reaction to
start with, and no offer.  If the reaction fails in the first phase (a
permanent failure, represented by \lstinline{Block}), the next attempt must be
made with an offer, to set up the blocking/signaling protocol.  If the
reaction fails in the second phase (a transient failure, represented by
\lstinline{Retry}), there is likely contention over shared data.  To reduce
the contention, it performs one cycle of exponential backoff before retrying.
If the reagent includes communication attempts, the retry is performed with an
offer, since doing so increases chances of elimination~(\secref{choice})
without further contention.  Finally, if both phases of the reaction succeed,
the final answer is returned.

The second retry loop, \lstinline{withOffer}, is similar, but begins by
allocating an \lstinline{Offer} object to make visible to other reagents.
%TODO: finish this

Initially, the reaction is attempted using \lstinline{withoutOffer},
representing optimism that the reaction can be completed without making a
visible offer.

\begin{figure}
\begin{lstlisting}[frame=single]
class Commit[A] extends Reagent[A,A] {
  def tryReact(a: A, rx: Reaction, offer: Offer[A]) = 
    if (offer != null && !offer.rescind) offer.answer
    else if (rx.commit) a 
    else Retry
}
\end{lstlisting}
\vskip -12pt
\begin{lstlisting}[frame=single]
class CAS[A](ref: Ref[A], oldV: A, newV: A) 
  extends Reagent[Unit,Unit] {
  def tryReact(u: Unit, rx: Reaction, offer: Offer[B]) = 
    if (!rx.hasCAS && !k.hasCAS)
      if (ref.cas(oldV, newV))
        k.tryReact((), rx, offer)
      else Retry
    else
      k.tryReact((), rx.withCAS(ref, oldV, newV), offer)
}
\end{lstlisting}
\vskip -12pt
\begin{lstlisting}[frame=single]
class Choice[A,B](r1: Reagent[A,B], r2: Reagent[A,B]) 
  extends Reagent[A,B] {
  def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
    r1.tryReact(a, rx, offer) match {
    case Retry => 
        r2.tryReact(a, rx, offer) match {
          case Retry => Retry
          case Block => Retry // must still retry r1
          case ans   => ans
        }
      case Block => r2.tryReact(a, rx, offer)
      case ans => ans
    }
}
\end{lstlisting}
\vskip -12pt
\begin{lstlisting}[frame=single]
class Computed[A,B](c: A => Reagent[Unit,B]) 
  extends Reagent[A,B] {
  def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
    c(a).tryReact((), rx, offer)
}
\end{lstlisting}
\nocaptionrule
\caption{Excerpts from core reagent subclasses}
\end{figure}

@@

As mentioned in~\secref{offers}, the initial continuation for reagents is the
\lstinline{Commit} continuation:

If the reagent has made an offer, but has also completed the first phase of
reaction, the offer must be rescinded before the commit phase is
attempted---otherwise, the reaction could complete twice.  As with the
\lstinline{!} method, the attempt to rescind the offer is in a race with other
reagents that may be completing the offer.  If \lstinline{Commit} loses the
race, it returns the answer provided by the offer.  Otherwise, it attempts to
commit the reaction, and if successful simply returns its input, which is the
final answer for the reaction.

Committing a reaction requires a $k$CAS operation: $k$ compare and sets must
be performed atomically.  This operation, which forms the basis of STM, is in
general expensive and not available through hardware.  There are several
software implementations that provide various progress
guarantees~\cite{Fraser2007,Luchangco2003,Attiya2008}, but they tend to

@@

@@

@@
